Light inextensible strings under tension in the Schwarzschild geometry

### 2019-08-28 string_physics_teacher.tex
work

General relativity is a topic of long-standing fascination for
undergraduate physics students~\cite{christensen2012}.  Here I show
how numerical methods, including computer algebra systems, can be used
to interpret the concept of ``straight line'' in the vicinity of a
black hole.  The analysis would be suitable for an undergraduate
homework problem~\cite{romano2019}.

@@

Further, because the system is symmetrical
under~$\phi\longrightarrow-\phi$, the string must be symmetrical about
at this point.  Considering~$r'(\phi)=0$ shows that a tangential
string at a point with~$r_0>2M$ obeys, to first order, $r''=r-2M$,
while tangential null geodesics follow $r''=r-3M$.  Thus a taut string
would {\em look} curved to a local observer, having a (visual) radius
of curvature of $r_0^2/M\sqrt{1-2M/r_0}$ (the extra term comes from
the fact that appearance is driven by local $ds$, not Schwarzschild
$r$).  At the photon sphere $r_0=3M$, such a string would appear to
have a radius of~$\sqrt{27}M$.  For a solar mass black hole this would
be about $15\unit{km}$, a readily understandable result.

@@

Here I solve equation~\ref{rdashdash} numerically, using the
\verb+deSolve+ package for ordinary differential
equations~\cite{soetart2010}.  Non-radial solutions fall into two
distinct classes: those with $r>2M$ at all points, and those which
attain $r=2M$, in which case the string is tangential to the event
horizon.

@@

The concept of ``straight line'' is problematic in general relativity.
Although photon paths provide a definition of straightness, their
world lines are sensitive to the time component of the metric.  Light
taut strings respond only to the spatial components of the metric, and
their configuration is arguably a more intuitive and helpful way to
think about straightness.  Non-trivial configurations are adopted by
such strings, including self-intersecting paths.  Taut strings are
arguably as fundamental a definition of straighn

Comparing light paths with that of nearby taut strings thus
furnishes another insight into path curvature, complementing the
analysis of Abramowicz~\cite{abramowicz1992}.

>>>>

General relativity is a topic of long-standing fascination for physics
undergraduates~\cite{christensen2012}.  Students appreciate everyday
illustrations of the concepts involved, as recent educational work on
gravitational wave visualisations attests~\cite{overduin2018}.  Here I
show how considering light inextensible string can be used to
interpret the concept of ``straight line'' in the vicinity of a black
hole.  The analysis presented here would be suitable for an
undergraduate homework problem~\cite{romano2019}.

@@

Further, because the system is invariant
under~$\phi\longrightarrow-\phi$, the string must be symmetrical about
at this point.  Considering~$r'(\phi)=0$ shows that a tangential
string at a point with~$r_0>2M$ obeys, to first order, $r''=r-2M$,
while tangential null geodesics follow $r''=r-3M$.  Thus a taut string
would {\em look} curved to a local observer, having a (visual) radius
of curvature of $r_0^2/M\sqrt{1-2M/r_0}$ (the extra term comes from
the fact that appearance is driven by local $ds$, not Schwarzschild
$r$).  At the photon sphere $r_0=3M$, such a string would appear to
have a radius of~$\sqrt{27}M$.  For a solar mass black hole this would
be about $15\unit{km}$, a readily understandable result.

@@

Here I solve equation~\ref{rdashdash} numerically, using R programming
language~\cite{rcore2019} and the \verb+deSolve+ package for ordinary
differential equations~\cite{soetart2010}.  Non-radial solutions fall
into two distinct classes: those with $r>2M$ at all points, and those
which attain $r=2M$, in which case the string is tangential to the
event horizon.

@@

The concept of ``straight line'' is problematic in general relativity.
Although photon paths provide a definition of straightness, their
world lines are sensitive to the time component of the metric.  Light
taut strings respond only to the spatial components of the metric, and
their configuration is arguably a more intuitive and helpful way to
think about straightness.  Using numerical methods to compute string
configuration is suitable for an undergraduate homework problem.
